\textbf{ Spectrophotometric parallaxes with \project{The Cygnet}:\\ Accurate distances for luminous red-giant stars }

### 2018-08-13 paper/sp.tex
noise model

% To-do items
% -----------
% - make sure we address the sensibleness and generality of the linear model for the photometric space
% - Coryn's point that we should carefully explain how our model is geometric.
% - related: HWR's point that this is standardizeable candles, like the Ia SNe results.
% - related: Put in an explanation of WHY this works so well.
% - related: Put in an explanation of how it is that we aren't really modeling luminosity or distance but really just generating parallaxes.
% - does the model need a cute name, like The Swan?
% - check with WISE experts that we have the correct WISE photometry
% - we need an uncertainty estimate or noise model for our outputs.
% - general HWR point that we aren't putting in to the model any idea about how apparent magnitude varies with distance; this means that we rely on data for EVERYTHING.
% - write words about the setting of Lambda and the parallax offset (currently 0.0483 +/- 0.012 mas)
% - we have shown that outliers in sp - gaia tend to have large H-W2 or excess flux like in crowded regions, but there are some exceptions. Our error estimates are consistent with no intrinsic scatter if we use a robust chi-squared estimate. That is, chi-squared is dominated by outliers. We clipped the bad apogee spectral pixels at an error of 0.05 in normalized flux. Our uncertainty estimates are sensitive to this. The reader is asked to suck it.
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?

@@

We need complete feature-vector information for all stars.  For this
reason, we define the Parent Sample to be the set of all stars that
meet the \apogee\ and \gaia\ cuts and also
have the complete set of photometry: $G_n$, $\BP_n$, $\RP_n$, $J_n$,
$H_n$, $K_n$, $W_{1n}$, and $W_{2n}$ and also an \apogee\ spectrum.
In addition, we applied two light color cuts to remove stars with obviously
contaminated or outlying photometry:
\begin{eqnarray}
(J_n - K_n) &<& (+0.4\,\mathrm{mag}) + 0.45 * (\BP_n - \RP_n)
\\
(H_n - W_{2n}) &>& (-0.05\,\mathrm{mag})
\quad .
\end{eqnarray}
These cuts removed roughly 2~percent of the \apogee\ stars.
This Parent Sample contains 44808 (HOGG CHECK) and is shown in \figurename~\ref{fig:parent}.
\begin{figure}
\includegraphics[width=\textwidth]{parent_sample.pdf}
\caption{Color--color diagrams for the Parent Sample stars.
  The points are colored by
  the \apogee\ pipeline estimates of surface gravity.
  The color cuts used to trim outliers are shown as grey lines.\label{fig:parent}}
\end{figure}

Every Parent-Sample star gets, in addition, a randomly assigned binary
label (A or B).
This is used for two-fold validation and jackknife.
In short, we will use the model trained on the A data to assign luminosities
and distances to the B data and \foreign{vice versa}.
The Parent Sample A has HOGG entries and Parent Sample B has HOGG entries.

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ DR2
the astrometric parallax $\gparallax_n$ and its uncertainty $\sigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset of $0.029$\,mas to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The Training Sets are shown in \figurename~\ref{fig:training}.
\begin{figure}
\includegraphics[width=\textwidth]{./training_sample.pdf}
\caption{Color--color diagrams for the Training Set stars.
  Note that there is less color range in the Training Set than in the
  Parent Sample.\label{fig:training}}
\end{figure}

@@

We adopt 9~percent as the method's precision.
This estimate is conservative in that the 9-percent scatter includes contributions
from both our spectrophotometric model and from the \gaia\ measurements themselves.
However this estimate is not so conservative in that the test is performed with
\gaia's best stars, which are bright, nearby, in low extinction regions, and
(because of these things) near Solar metallicity.
There could in principle be additional bias or scatter for stars that are in dustier
regions or at lower metallicities.
\figurename~\ref{fig:residuals} shows that
there is no suggestion of any such trends when we color the residuals by metallicity
or $H-W_2$ color (which is a reddening proxy).
\begin{figure}
\includegraphics[width=\textwidth]{./residuals_training.pdf}
\caption{Prediction residuals colored by relevant features.
  Each panel shows the prediction residuals (\gaia\ astrometric parallax minus
  spectrophotometric parallax $\gparallax_n-\sparallax_n$) colored by a different
  data feature that might be relevant to the residual.\label{fig:residuals}}
\end{figure}

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from this work for stars in
angular proximity to known stellar clusters and the Sagittarius dwarf
galaxy.
These clusters span some range in metallicity and age, and therefore test
the accuracy of the method for different kinds of populations.
They show that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
HOGG: DO WE NEED TO ADDRESS whether why not these clusters show 9-percent spreads?
\begin{figure}
\includegraphics[width=\textwidth]{clusters.pdf}
\caption{The histograms of \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for stars in angular proximity
  to various stellar clusters in the \apogee\ observing footprint. In each panel,
  the expected parallax of the cluster is shown as a vertical line.\label{fig:clusters}}
\end{figure}

@@

In order to use these outputs in a probabilistic inference or forward model,
an investigator needs a noise (uncertainty) model.
The linear model permits straightforward propagation of uncertainty from the
feature inputs.
However, these propagated uncertainties clearly under-predict the scatter between
the spectrophotometric parallaxes $\sparallax_m$ and the \gaia\ astrometric
parallaxes $\gparallax_m$.
Our recommendation for an uncertainty model is HOGG WAT?

@@

One significant limitation of the model is that it does not produce a sensible or
useful noise estimate or error bar on each spectrophotometric parallax estimate.
Although the linear model does permit simple noise propagation, that noise propagation
seriously under-estimates the scatter between the spectrophotometric parallaxes
and the best of the \gaia\ \acronym{DR2} astrometric parallaxes.
We thus used an empirical noise estimate to get our HOGG DETAIL HERE.
In principle it might have been better to build a noise model for our outputs,
and learned that noise model along with the coefficients.

>>>>

% To-do items
% -----------
% - make sure we address the sensibleness and generality of the linear model for the photometric space
% - Coryn's point that we should carefully explain how our model is geometric.
% - related: HWR's point that this is standardizeable candles, like the Ia SNe results.
% - related: Put in an explanation of WHY this works so well.
% - related: Put in an explanation of how it is that we aren't really modeling luminosity or distance but really just generating parallaxes.
% - does the model need a cute name, like The Swan?
% - check with WISE experts that we have the correct WISE photometry
% - general HWR point that we aren't putting in to the model any idea about how apparent magnitude varies with distance; this means that we rely on data for EVERYTHING.
% - write words about the setting of Lambda and the parallax offset (currently 0.0483 +/- 0.012 mas)
% - we have shown that outliers in sp - gaia tend to have large H-W2 or excess flux like in crowded regions, but there are some exceptions. Our error estimates are consistent with no intrinsic scatter if we use a robust chi-squared estimate. That is, chi-squared is dominated by outliers. We clipped the bad apogee spectral pixels at an error of 0.05 in normalized flux. Our uncertainty estimates are sensitive to this. The reader is asked to suck it.
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?

@@

We need complete feature-vector information for all stars.  For this
reason, we define the Parent Sample to be the set of all stars that
meet the \apogee\ and \gaia\ cuts and also
have the complete set of photometry: $G_n$, $\BP_n$, $\RP_n$, $J_n$,
$H_n$, $K_n$, $W_{1n}$, and $W_{2n}$ and also an \apogee\ spectrum.
In addition, we applied two light color cuts to remove stars with obviously
contaminated or outlying photometry:
\begin{eqnarray}
(J_n - K_n) &<& (+0.4\,\mathrm{mag}) + 0.45 * (\BP_n - \RP_n)
\\
(H_n - W_{2n}) &>& (-0.05\,\mathrm{mag})
\quad .
\end{eqnarray}
These cuts removed roughly 2~percent of the \apogee\ stars.
This Parent Sample contains 44808 stars and is shown in \figurename~\ref{fig:parent}.
\begin{figure}
\includegraphics[width=\textwidth]{parent_sample.pdf}
\caption{Color--color diagrams for the Parent Sample stars.
  The points are colored by
  the \apogee\ pipeline estimates of surface gravity.
  The color cuts used to trim outliers are shown as grey lines.\label{fig:parent}}
\end{figure}

Every Parent-Sample star gets, in addition, a randomly assigned binary
label (A or B).
This is used for two-fold validation and jackknife.
In short, we will use the model trained on the A data to assign luminosities
and distances to the B data and \foreign{vice versa}.

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\sigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value of $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt a value of $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation likelihood.
The spectrophotometric parallaxes do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in \figurename~\ref{fig:training}.
\begin{figure}
\includegraphics[width=\textwidth]{./training_sample.pdf}
\caption{Color--color diagrams for the Training Set stars.
  Note that there is less color range in the Training Set than in the
  Parent Sample.\label{fig:training}}
\end{figure}

@@

We adopt 9~percent as the method's precision.
This estimate is conservative in that the 9-percent scatter includes contributions
from both our spectrophotometric model and from the \gaia\ measurements themselves.
However this estimate is not so conservative in that the test is performed with
\gaia's best stars, which are bright, nearby, in low extinction regions, and
(because of these things) near Solar metallicity.
There could in principle be additional bias or scatter for stars that are in dustier
regions or at lower metallicities.
\figurename~\ref{fig:residuals} shows that
there is no suggestion of any such trends when we color the residuals by metallicity
or $H-W_2$ color (which is a reddening proxy).
However, that figure does show that stars with greater reddening do appear to have
larger scatter against the \gaia\ parallaxes.
Additional tests confirm that our spectroscopic parallaxes have largest scatter
in the dustiest and most crowded regions, as expected given the relatively low angular resolution
photometry.
\begin{figure}
\includegraphics[width=\textwidth]{./residuals_training.pdf}
\caption{Prediction residuals colored by relevant features.
  Each panel shows the prediction residuals (\gaia\ astrometric parallax minus
  spectrophotometric parallax $\gparallax_n-\sparallax_n$) colored by a different
  data feature that might be relevant to the residual.\label{fig:residuals}}
\end{figure}

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from this work for stars that are
astrometrically confirmed members of known stellar clusters.
These clusters span some range in metallicity and age, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, these figures 
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlates
with parallax offsets.
\begin{figure}
\includegraphics[width=\textwidth]{clusters.pdf}
\caption{The histograms of \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  four stellar clusters in the \apogee\ observing footprint. The spectroscopic
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  the expected parallax of the cluster is shown as a vertical line.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

The linear model permits straightforward propagation of uncertainty from the
feature inputs to the spectrophotometric parallaxes.
Although the model fundamentally assumes that the prediction features are noise-free
(see the ``good features'' assumption in \sectionname~\ref{sec:assumptions}),
there are in fact small uncertainties on the features that we can propagate to 
make uncertainty estimates on the spectroscopic parallaxes.
We perform this propagation; these uncertainties are also given in
\tablename~\ref{tab:output}.

@@

HOGG SOMETHING ABOUT THE CLUSTER M71 BEING CONCERNING AND OUR EXPLANATIONS.
